---
title: "Project 2: Fine particulate matter (PM2.5)"
output: pdf_document
---

```{r include=FALSE}
NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")
```

```{r}
library(tidyverse)
library(scales)
```
## 0 Introduction
### 0.1 PM2.5 Emissions Data (NEI): 
all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year.

Variables:\newline
- fips: A five-digit number (represented as a string) indicating the U.S. county \newline
- SCC: The name of the source as indicated by a digit string (see source code classification table)\newline
- Pollutant: A string indicating the pollutant\newline
- Emissions: Amount of PM2.5 emitted, in tons\newline
- type: The type of source (point, non-point, on-road, or non-road)\newline
- year: The year of emissions recorded\newline

```{r}
head(NEI)
```

### 0.2 Source Classification Code Table (SCC):
a mapping from the SCC digit strings in the Emissions table to the actual name of the PM2.5 source. The sources are categorized in a few different ways from more general to more specific and you may choose to explore whatever categories you think are most useful.

```{r}
head(SCC)
```

## 1 

Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? 

Yes, total emissions have decreased.
```{r}
year=c(1999,2002,2005,2008)
total_emissions=c()

for (i in 1:4){
  NEI_year=NEI[NEI$year==year[i],]
  total_emissions[i]=sum(NEI_year$Emissions)
  assign(paste0("NEI_year_",year[i]),NEI_year)
}

png(filename="plot1.png")
barplot(total_emissions,names = year, xlab = "Years", ylab = "Tons", 
        main = "Total PM2.5 emissions in the United States")
dev.off()
```

## 2

Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips==24510) from 1999 to 2008? 

Yes, total emissions in Baltimore have decreased.
```{r}
total_emissions_Baltimore=c()

for (i in 1:4){
  NEI_year=get(paste0("NEI_year_",year[i]))
  NEI_year_Baltimore=NEI_year[NEI_year$fips=="24510",]
  total_emissions_Baltimore[i]=sum(NEI_year_Baltimore$Emissions)
}

png(filename="plot2.png")
barplot(total_emissions_Baltimore,names = year, xlab = "Years", ylab = "Tons"
        , main = "Total PM2.5 emissions in Baltimore City, 1999-2008")
dev.off()
```

## 3

Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008?

Onroad, Non-road and Nonpoint have seen decreases in emissions. Point has seen decreases in emissions.

```{r}
png(filename="plot3.png")
NEI%>%filter(fips=="24510")%>%
  select(Emissions,type,year)%>%
  mutate(type=factor(type,levels = c("ON-ROAD","NON-ROAD","POINT","NONPOINT")),
         year=factor(year))%>%
  group_by(type,year)%>%
  summarise(total=sum(Emissions))%>%#ungroup()%>%
  ggplot(aes(x=year,y=total))+
  geom_bar(stat="identity")+
  facet_wrap(~type,ncol=4)+
  labs(title = "Total PM2.5 emissions by types in Baltimore City",
       y="Tons")+
  theme(plot.title = element_text(hjust = 0.5))
dev.off()
```

## 4 
Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?

Emissions from coal combustion-related sources have decreased.
```{r}
SCC_Comb.Coal=SCC%>%
  filter(str_detect(Short.Name,"Comb|Coal"))%>%
  pull(SCC)

png(filename="plot4.png")
NEI%>%filter(SCC %in% SCC_Comb.Coal)%>%
  select(Emissions,year)%>%
  mutate(year=factor(year))%>%
  group_by(year)%>%
  summarise(total=sum(Emissions))%>%
  ggplot(aes(x=year,y=total))+
  geom_bar(stat="identity")+
  labs(title = "Total PM2.5 emissions from coal combustion-related sources",
       y="Tons")+
  theme(plot.title = element_text(hjust = 0.5))
dev.off()
```

## 5

How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?

Emissions from motor vehicle sources have decreased.
```{r}
SCC_vehicle=SCC%>%
  filter(str_detect(SCC.Level.Two,"Vehicle"))%>%
  pull(SCC)

png(filename="plot5.png")
NEI%>%filter(fips=="24510" & SCC %in% SCC_vehicle)%>%
  select(Emissions,year)%>%
  mutate(year=factor(year))%>%
  group_by(year)%>%
  summarise(total=sum(Emissions))%>%
  ggplot(aes(x=year,y=total))+
  geom_bar(stat="identity")+
  labs(title = "Total PM2.5 emissions from motor vehicle sources in Baltimore City",
       y="Tons")+
  theme(plot.title = element_text(hjust = 0.5))
dev.off()
```

## 6 

Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips=="06037"). Which city has seen greater changes over time in motor vehicle emissions?

Baltimore City has seen greater changes in motor vehicle emissions, we can tell from dividing emissions in later years by emissions in 1999.
```{r}
(NEI_vehicle_city=NEI%>%
  filter(fips %in% c("24510","06037") & SCC %in% SCC_vehicle)%>%
  dplyr::select(Emissions,fips,year)%>%
  mutate(fips=factor(fips,levels=c("24510","06037"),labels=c("Baltimore City","Los Angeles")),
         year=factor(year))%>%
  group_by(fips,year)%>%
  summarise(total=sum(Emissions))%>%
  ungroup()%>%group_by(fips)%>%
  mutate(percent_year1999=percent(total/total[1])))

png(filename="plot6.png")
NEI_vehicle_city%>%ggplot(aes(x=year,y=total))+
  geom_bar(stat="identity")+
  facet_wrap(~fips,ncol=2)+
  labs(title = "Total PM2.5 emissions from motor vehicle sources in Baltimore and Los Angeles",
       y="Tons")+
  theme(plot.title = element_text(hjust = 0.5))
dev.off()
```

